Inferring neural networks from fluorescent calcium imaging using partial correlation and sample weighting

### 2014-05-22 article.tex
add todo
problem definition + partial correlation + filtering

@@

 % Three or more authors with the same address:
 \author{}

@@

\begin{abstract}
This is the abstract for this article.
\end{abstract}

@@

% BG : brain, complex organ

Understanding how the brain works is a key to understand and to treat
brain pathologies and disorders such as Parkinson's disease, epilepsy.
Retrieving the connectomes, neurons connectivity map, will shed new lights on
the anatomical and functional connectivity of the brain.
The human brain is a complex biological organ, which is formed by 100
billions of neurons with 7000 synaptic connections on average. Inferring the
neuron connectivity network through dissection would be an impossible
and daunting task.

% Issue : we have data, but need to have to infer the connectome
Cutting edge optical devices (TODO CITE) allows to track simultaneously
the neural activity of thousands of neurons through the fluorescent
calcium indicator. The neuron connectivity network could be inferred
from those neuron activity signals.

% Solution
In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation to infer the neural connectivity network.
This is also the winning method of the Connectomics challenge.

@@

\section{Empirical experiments}
\subsection{Datasets}
\subsection{Metrics}
\subsection{Data filtering}
\subsection{Signal distribution}
\subsection{ROC AUC results}
\subsection{AUPRC}

>>>>

%% Personnal package
\usepackage{enumerate}

@@

 % Three or more authors with the same address:
 \author{Todo authors}

@@

\begin{abstract}
Understanding how the brain works is a key to understand and to treat
brain pathologies and disorders such as Parkinson's disease, epilepsy.
Retrieving the connectomes, neurons connectivity map, will shed new lights on
the anatomical and functional connectivity of the brain.
The connectomics challenge

\end{abstract}

@@

% BG : brain, complex organ
Understanding how the brain works is a key to understand and to treat
brain pathologies and disorders such as Parkinson's disease, epilepsy.
Retrieving the connectomes, neurons connectivity map, will shed new lights on
the anatomical and functional connectivity of the brain.
The human brain is a complex biological organ, which is formed by 100
billions of neurons with 7000 synaptic connections on average. Inferring the
neuron connectivity network through dissection would be an impossible
and daunting task.

% Issue : we have data, but need to have to infer the connectome
% TODO Shed light on the difficulty of the inverse problem:
% noise, time scale (optical imaging speed vs neuron speed), decay of
% fluorescent signal.
Cutting edge optical devices [todo cite] allows to track simultaneously
the neural activity of thousands of neurons through the fluorescent
calcium indicator. The neuron connectivity network could be inferred
from those neuron activity signals.

The neural connectivity inference consists in highlighting, from observational data, direct relationships between neurons. One way to emphasize the connections between neurons is to represent them by a (un)directed graph with $p$ nodes, where each node represents a neuron, and an edge $e_{ij}$ which represents a direct link from neuron  $i$ to neuron $j$. If the graph is undirected, then $e_{ij}$ is equal to $e_{ji}$  for all $i,j$.

% Solution
In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation to infer the neural connectivity network.
This is also the winning method of the Connectomics challenge [todo cite].

@@

\paragraph{Problem definition\\}

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
Formally, the challenge problem may be formulated by the following statement : 
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of fluorescent
calcium concentration of size $n$ from $p$ neurons, the goal is to infer the
connectivity network $y \in \left\{0, 1\right\}^{p \times p}$ where
non zero values indicate that the neuron $i$ have an outward connection to
the neuron $j$.}

The challenge was stated as an unsupervised learning problem. The particularity of unsupervised inference is the network recovery solely based on observational data. Other approaches have been considered to infer biochemical networks, e.g. gene regulatory network, such as supervised learning [todo cite] or transfer learning [todo cite], but these learning processes require, as starting point, a partial knowledge of the target network or known networks similar to the unknown one.

% Introduce partial correlation
\paragraph{Partial correlation\\}

A common association measure is the Pearson correlation coefficient. 
\begin{align}
\rho_{ij} = \dfrac{cov(I,J)}{\sigma_I \sigma_J}
\end{align}
Even if \textit{correlation does not imply causation} [todo cite], it may give an insight on the underlying connectivity network. Nevertheless, the main challenge in inferring a network is to distinguish direct from indirect interactions [de la Fuente et al., 2004]. While the classical correlation coefficient characterize both direct and indirect interactions, the partial correlation coefficient quantifies association between variables conditionally on one or several other variables (the number of variables considered being the order of the partial correlation coefficient). [Shipley, 2002] suggests to consider correlation coefficients of all possible orders, i.e. conditioned on every subsets of the set of $p-2$ other variables, in order to take into account multiple indirect paths between a given pair of variables.  An obvious limitation to calculate high-order coefficient is the number of samples in conditioned data sets. Considering the large number of samples at our disposal in the connectomics challenge, we only consider the highest order of partial correlation coefficient. We thus try to estimate partial correlation of order $p-2$ $\rho_{ij | \mathcal{V}^{-i,j}}$. A way to find these coefficient [todo cite?] is to take the inverse of the covariance matrix to get the partial correlation matrix of order $p-2$, also known as the precision matrix [todo cite] .

% Introduce directivity
\paragraph{Directivity}

% Introduce filtering
\paragraph{Filtering\\}

The filtering process is made of three main parts:

\begin{enumerate}[i.]
\item A low-pass filter: we consider four shapes of local averaging, 
\item A high pass filter : we apply the asymmetric discrete derivative,
\item A non-linear filter: we re-weight samples depending on the current activity of the systems in order to favour the association measure when directly connected neurons are firing together, while artificially reducing the association measure in case of extreme network activity, i.e. high when a network firing is happening and low when no neuron is active.
\end{enumerate}

\section{Empirical experiments}
\subsection{Datasets}

\paragraph{Improvement of each stage of our approach\\}
\paragraph{Comparison with other methods\\}

\subsection{Metrics}
\subsection{Data filtering}
\subsection{Signal distribution}
\subsection{ROC AUC results}
\subsection{AUPRC}
